Applying Radial Basis Function Networks and Markov Chains for on-line detection of concept drift in non-stationary environments

### 2019-10-28 tex/article/article.tex
concept drift iter 1

A significant portion of this data is produced in the form of uninterrupted and potentially infinite sequences \cite{Aggarwal:2006:DSM:1196418}.
In the literature,
sequences with these characteristics are called data streams.
These streams are present in various fields of application,
such as financial market monitoring  \cite{ZHOU:2015},
road traffic monitoring \cite{Wang:2015:EOV:2843092.2843464},
telecom network management \cite{delattre2015method},
real-time sentiment analysis  \cite{KRANJC2015187}
and intruder prevention and identification systems \cite{KENKRE:PAI:COLACO:2015}.

@@

The rest of the paper is organized as follows:
Section 2 describes the concept drift phenomenon and the main detection techniques;
Section 3 presents the eye-tracking problem;
Section 4 describes the RBFChain algorithm and its pseudo-code;
Section 5 shows the configuration and results for the experiment with synthetic datasets;
Section 6 presents the configuration and results for the experiment with the eye-tracking problem; and, finally, Section 7 provides conclusions and discusses future work.

@@

Most real-world problems can be regarded as non-stationary environments \cite{Gama:2014:DAF:2670967.2670971}.
In these environments, the joint probability distribution can change over time,
such as a switch in the conditional probability distribution on a classification problem,
or a change of some moment (such as mean and variance) on a time series forecasting problem \cite{tsymbal2004problem}.
%
Systems applied to these environments may be unable to adapt to the new information,
hence dramatically deteriorating its performance.
This phenomenon is called concept drift.

The Bayesian Theory can be used as a background to formally define the concept drift phenomenon \cite{Elwell:2011}:
consider the posterior probability of a sample $x$ belonging to a class $y$, a concept drift happens when this probability changes over time, that is, $Pt + 1 (y | x) \neq Pt (y | x)$. In a supervised learning scenario, this can be interpreted as when the relationship between the input data and the target variable change over time.

According to \cite{tsymbal2004problem, Gama:2014:DAF:2670967.2670971}, concept drifts can occur in four main patterns:

\begin{itemize}
    \item \textbf{Abrupt:} occurs when a concept A switches abruptly to another concept B.
    \item \textbf{Gradual:} occurs when a concept A is being exchanged for the B concept gradually. In this case, while there is no definitive change from concept A to concept B, occurrences of B become more frequent, while fewer events of A are observed.
    \item \textbf{Incremental:} occurs when a concept A is being exchanged for B through intermediate concepts.  These concepts differ little from its predecessor and successor, so changes are noticeable only in the long run.
    \item \textbf{Recurrent:} occurs when a previously active concept reappears after a certain period. However, this can not be understood as a periodic seasonality.
\end{itemize}

Figure \ref{fig:concept_drift_patterns} demonstrates these patterns:

@@

Algorithms for detecting concept drift characterize and quantify concept drifts through the delimitation of the moments or time intervals in which changes occur \cite{Basseville:1993:DAC:151741}.
%
These algorithms fall into two categories, according to the need for data labeling \cite{Zliobaite:2010}:

\begin{itemize}
    \item \textbf{Explicit Algorithms/Supervised}: Adopt a passive approach, as they depend on the correct labeling of the data.
    The model performance is monitored continuously, and drifts are detected when its performance starts to deteriorate, passing past a threshold.
    \item \textbf{Implicit Algorithms/Unsupervised:} Implement a proactive approach and are independent of correct data labeling.
    Concept drifts are detected through the analysis of incoming data or indicators produced by the applied learning techniques.
    Although they are more prone to false alarms, they are an alternative to scenarios where obtaining labels is expensive, time-consuming or unviable.
    Also, this approach can lead to better results, since it is possible to refit the model or adjust the data before the deterioration of the predictions.
\end{itemize}

The algorithm proposed in this paper, RBFChain, classifies as an unsupervised algorithm with a proactive approach.

>>>>

A significant share of this data is produced in the form of uninterrupted and potentially infinite sequences \cite{Aggarwal:2006:DSM:1196418}.
In the literature,
sequences with these characteristics are known as data streams.
These streams are present in various fields of application,
such as financial market monitoring  \cite{ZHOU:2015},
road traffic monitoring \cite{Wang:2015:EOV:2843092.2843464},
telecom network management \cite{delattre2015method},
real-time sentiment analysis  \cite{KRANJC2015187}
and intruder prevention and identification systems \cite{KENKRE:PAI:COLACO:2015}.

@@

The rest of the paper is organized as follows:
Section 2 describes the concept drift phenomenon and the main detection techniques;
Section 3 presents the eye-tracking problem;
Section 4 describes the RBFChain algorithm and its pseudo-code;
Section 5 presents the setup and results of the experiment with synthetic datasets;
Section 6 shows the setup and results of the eye-tracking experiment;
and, finally, Section 7 provides conclusions and discusses future work.

@@

Most real-world problems scenarios can be regarded as non-stationary environments \cite{Gama:2014:DAF:2670967.2670971}.
In these environments, the joint probability distribution can change over time,
such as a switch in the conditional probability distribution on a classification problem,
or a change of some moment (such as mean and variance) on a time series forecasting problem \cite{tsymbal2004problem}.
%
Systems applied to these settings may be unable to adapt to the changes, hence dramatically deteriorating their performance.
This phenomenon is called concept drift.

The Bayesian Theory can be used to formally define the concept drift phenomenon \cite{Elwell:2011}:
consider the posterior probability of a sample $x$ belonging to a class $y$, a concept drift happens when this probability changes over time, that is, $Pt + 1 (y | x) \neq Pt (y | x)$. In a supervised learning scenario, this can be interpreted as when the relationship between the input data and the target variable change over time.

According to \cite{tsymbal2004problem, Gama:2014:DAF:2670967.2670971}, concept drifts can occur in four main patterns.
These patterns are demonstrated in Figure \ref{fig:concept_drift_patterns} and described below:

\begin{itemize}
    \item \textbf{Abrupt:} occurs when a concept A switches abruptly to another concept B.
    \item \textbf{Gradual:} occurs when a concept A is being exchanged for the B concept slowly. In this case, while there is no definitive change from concept A to concept B, occurrences of B become more frequent, while fewer events of A are observed.
    \item \textbf{Incremental:} occurs when a concept A is being exchanged for B through intermediate concepts.  These concepts differ little from its predecessor and successor, so changes are noticeable only in the long run.
    \item \textbf{Recurrent:} occurs when a previously active concept reappears after a certain period. However, this cannot be understood as a periodic seasonality.
\end{itemize}

@@

Concept drift detection methods characterize and quantify concept drifts through the delimitation of moments or time intervals in which change happens \cite{Basseville:1993:DAC:151741}.
%
These algorithms fall into two categories, according to the need for data labeling \cite{Zliobaite:2010}:

\begin{itemize}
    \item \textbf{Explicit Algorithms/Supervised}: Adopt a reactive approach, as they depend on the correct labeling of the data.
    The model performance is monitored continuously, and drifts are detected when its performance starts to deteriorate, passing past a threshold.
    \item \textbf{Implicit Algorithms/Unsupervised:} Implement a proactive approach and are independent of correct data labeling.
    Concept drifts are detected through the analysis of incoming data or indicators produced by the applied learning techniques.
    Although more prone to false alarms, they are an alternative to scenarios where obtaining labels is expensive, time-consuming, or unviable.
    Also, this approach can lead to better results, since it is possible to refit the model or adjust the data before the deterioration of the predictions.
\end{itemize}

This paper proposes a novel unsupervised and proactive concept drift detection method.
The algorithm continually groups the incoming data through Radial Basis Function Networks and maintains a model of the drift transitions in a Markov Chain.
Drifts are detected when the probability of a transition reaches a parametric threshold.
